Model Checking Differentially Private Properties

### 2018-01-28 applications.tex
add privacy analysis and specification

Consider the MDP in Figure~\ref{figure:simple-mdp-mdp}. Suppose the
states $+$ and $-$ are neighbors. Let us compute $\{ s : M,
\neighbor{S}, s \models \dpriv{3}{0} (\F Y) \}$. First, the self
product $M^2$ is depicted in Figure~\ref{figure:simple-mdp-product}
(the states $++$ and $--$ and associated transitions are ignored
for clarity). Applying
Theorem~\ref{theorem:eventuallyC} with $\epsilon = 4$ and $\delta = 0$, 
consider the following linear programming problem:
\[
\begin{array}{rcrcrcrcrcl}
  \multicolumn{11}{c}{
  \min x_{++} + x_{+-} + x_{-+} + x_{--} + x_{YY} + x_{YN} + x_{NY} + x_{NN}
  }\\
  \multicolumn{9}{r}{ x_{YN} } & = & 1 \\
  \multicolumn{9}{r}{ x_{YY} } & = & -3 \\
  \multicolumn{9}{r}{ x_{NY} } & = & -4 \\
  \multicolumn{9}{r}{ x_{NN} } & = & 0 \\
  x_{++} &-&   .5625 x_{YY} &-& .1875 x_{YN} &-& .1875 x_{NY} &-&
               .0625 x_{NN} & \geq & 0\\
  x_{++} &-&   .64   x_{YY} &-& .16   x_{YN} &-& .16   x_{NY} &-&
               .04   x_{NN} & \geq & 0\\
  x_{+-} &-&   .1875 x_{YY} &-& .5625 x_{YN} &-& .0625 x_{NY} &-& 
               .1875 x_{NN} & \geq & 0\\
  x_{+-} &-&   .16   x_{YY} &-& .64   x_{YN} &-& .04   x_{NY} &-&
               .16   x_{NN} & \geq & 0\\
  x_{-+} &-&   .1875 x_{YY} &-& .0625 x_{YN} &-& .5625 x_{NY} &-& 
               .1875 x_{NN} & \geq & 0\\
  x_{-+} &-&   .16   x_{YY} &-& .04   x_{YN} &-& .64   x_{NY} &-&
               .16   x_{NN} & \geq & 0\\
  x_{--} &-&   .0625 x_{YY} &-& .1875 x_{YN} &-& .1875 x_{NY} &-&
               .5625 x_{NN} & \geq & 0\\
  x_{--} &-&   .04   x_{YY} &-& .16   x_{YN} &-& .16   x_{NY} &-&
               .64   x_{NN} & \geq & 0
\end{array}
\]
The optimum is attained at 
$(x_{++}, x_{+-}, x_{-+}, x_{--}) = (-2.25, 0, -2.75, -0.6)$. By
Theorem~\ref{theorem:eventuallyC}, we have
\[
\begin{array}{l}
\max_{\scheduler{S}}
\myPr{+}{M_{\scheduler{S}}}{\neighbor{S}}{\F Y} -
4 \myPr{+}{M_{\scheduler{S}}}{\neighbor{S}}{\F Y}
= -2.25 \leq 0\\
\max_{\scheduler{S}}
\myPr{+}{M_{\scheduler{S}}}{\neighbor{S}}{\F Y} -
4 \myPr{-}{M_{\scheduler{S}}}{\neighbor{S}}{\F Y}
= 0 \leq 0 \\
\max_{\scheduler{S}}
\myPr{-}{M_{\scheduler{S}}}{\neighbor{S}}{\F Y} -
4 \myPr{+}{M_{\scheduler{S}}}{\neighbor{S}}{\F Y}
= -2.75 \leq 0\\
\max_{\scheduler{S}}
\myPr{-}{M_{\scheduler{S}}}{\neighbor{S}}{\F Y} -
4 \myPr{-}{M_{\scheduler{S}}}{\neighbor{S}}{\F Y}
= -0.6 \leq 0
\end{array}
\]
Therefore $+, - \in \{ s : M, \neighbor{M}, s \models \dpriv{3}{0}
(\F Y) \}$. This is expected. Recall that low- and high-accuracy
mechanisms are $(3, 0)$- and $(4, 0)$-differentially private
respectively. Regardless of request types, the interactive mechanism
in Figure~\ref{figure:simple-mdp-mdp} can only achieve $(4,
0)$-differential privacy at best.

Streaming processing gives another motivating example for Markov
decision processes. In stream processing, input data are retrieved and
analyzed continuously. Such continual observation is commonly found
in social media, service providers, or even electronic surveillance
systems. To protect privacy, privacy-respecting mechanisms must be
formulated and designed. Classical differential privacy however is
insufficient in streaming processing. Note that observations are of
indefinite lengths in such applications. All differentially private
mechanism necessarily stay in their internal states indefinitely
before outcomes are produced. Although differential privacy demands
privacy-respecting outcomes, internal states may reveal private
information in such mechanisms. When differentially private mechanisms
are in their internal states, privacy is still lost if an intrusion
occurs.

@@

Based on the density estimation algorithm in \cite{DNPRY:10:PPSA}, we
consider the following mechanism for event type streams. Our mechanism
keeps a flag for each event type.

@@

For each event, a bit denotes whether it has occurred or not. To model
adjacent inputs, we use four actions symbols to denote events and its
replacement. For instance, the action $a[a]$ means the event $a$ is
observed without replacement; $b[a]$ means the event $b$ is observed
but replaced with $a$ in an adjacent input.

@@

      (01) edge [loop below] node { $a[\uscore]$, .375 } (01)
      (01) edge [bend left=10] node { $a[\uscore]$, .675 } (11)
      (01) edge [loop left] node [above=1,rotate=90] { $b[\uscore]$, .625 } (01)
      (01) edge [bend left=10] node [left=-3,above=1,rotate=90] { $b[\uscore]$, .375 } (00)

@@

      (01') edge [loop below] node { $\uscore[a]$, .375 } (01')
      (01') edge [bend left=10] node { $\uscore[a]$, .675 } (11')
      (01') edge [loop left] node [above=1,rotate=90] { $\uscore[b]$, .625 } (01')
      (01') edge [bend left=10] node [left=-3,above=1,rotate=90] { $\uscore[b]$, .375 } (00')

@@

\todo{verify this is $1$-DP}

>>>>

Streaming processing gives a motivating example for Markov
decision processes. In stream processing, input data are retrieved and
analyzed continuously. Such continual observation is commonly found
in social media, service providers, or even electronic surveillance
systems. To protect privacy, privacy-respecting mechanisms must be
formulated and designed. Classical differential privacy however is
insufficient in streaming processing. Note that observations are of
indefinite lengths in such applications. All differentially private
mechanism necessarily stay in their internal states indefinitely
before outcomes are produced. Although differential privacy demands
privacy-respecting outcomes, internal states may reveal private
information in such mechanisms. When differentially private mechanisms
are in their internal states, privacy is still lost if an intrusion
occurs.

@@

Based on the density estimation algorithm in \cite{DNPRY:10:PPSA}, we
consider the following mechanism for event type streams. Our mechanism
keeps a bit for each event type. The bits are initialized to $0$ or
$1$ uniformly. On observing an event type in the input stream, the
bit for the event type is set to $0$ and $1$ with probability
$\frac{3}{8}$ and $\frac{5}{8}$ respectively. The mechanism continues
to update its bits until it is asked to return the number of observed
event types. To compute the number of observed event types, the
mechanism adopts the truncated $1/2$-geometric mechanism on the sum of
two bits for event types. If the sum is $0$, the mechanism outputs $0$
with probability $\frac{2}{3}$ and $1, 2$ each with probability
$\frac{1}{6}$. If the sum is $1$, one of $0, 1, 2$ is returned with
the uniform distribution. If the sum is $2$, the mechanism outputs $2$
with probability $\frac{2}{3}$ and $0, 1$ each with probability
$\frac{1}{6}$ (Algorithm~\ref{algorithm:frequency-estimator}).

@@

\indent
\emph{Privacy analysis}.
In the processing phase, the bit $f_a$ for an event type $a$ can only
have two different distributions $U$ and $B$. If the event type is
never observed, $\Pr_U[f_a = 0] = \Pr_U[f_a = 1] = \frac{1}{2}$; if
the event type has been observed, we have $\Pr_B[f_a = 0] =
\frac{3}{8}$ and $\Pr_B[f_a = 1] = \frac{5}{8}$. Note that
$\frac{1}{2} \cdot \frac{3}{8} \leq \frac{1}{2} \leq 2 \cdot
\frac{3}{8}$ and $\frac{1}{2} \cdot \frac{5}{8} \leq \frac{1}{2} \leq
2 \cdot \frac{5}{8}$. For every $x \in \{ 0, 1 \}$, we have
$\frac{1}{2} \cdot \Pr_B[f_a = x] \leq \Pr_U[f_a = x] \leq 2 \cdot
\Pr_B[f_a = x]$. On neighboring observations, the distribution of
$f_a$ can only be changed from $U$ to $B$, values of $f_a$ is $(2,
0)$-differentially private and similarly for $f_b$. Finally, the
$1/2$-geometric mechanism gives $(1/2, 0)$-differential privacy. By
composition theorem, the steaming data analysis mechanism in
Algorithm~\ref{algorithm:frequency-estimator} is $(1,
0)$-differentially private.

To model Algorithm~\ref{algorithm:frequency-estimator}, let
us start with the formalization of observations. Note that the
lengths of observations are finite but unbounded. An infinite Markov
chain would be necessary if observations were formalized as states. 
Recall that neighbors are observations of the same length. If we
furthermore consider the formalization of neighbors, it is not clear
how to construct a finite Markov chain for
Algorithm~\ref{algorithm:frequency-estimator}.

In order to obtain a finite model, we formalize neighbors as action
sequences in Markov decision processes. Consider the action set $\{
a[a], a[b], b[a], b[b] \}$. Informally, the action $x[y]$ means the
event type $x$ is replaced by the event type $y$ in a neighbor. An
action sequence thus represents an observation and one of its
neighbors and vice versa. More formally, the action sequence
$x_1[y_1]x_2[y_2]\cdots x_n[y_n]$ represents the observation
$x_1x_2\cdots x_n$ and its neighbor $y_1y_2\cdots y_n$. For instance,
the action sequence $a[a]a[b]a[a]b[a]b[b]$ represents the observation
$aaabb$ and a neighbor $abaab$.

Similar to Algorithm~\ref{algorithm:frequency-estimator}, our Markov
decision process has internal states $00$, $01$, $10$, $11$ for the
two bits $f_a$ and $f_b$ of observed event types
(Figure~\ref{figure:adjacent-inputs}). The internal state $01$
represents $(f_a, f_b) = (0, 1)$, for instance. From the
initialization state $I$, the Markov decision process moves to one of
the four internal states randomly with the uniform
distribution. Consider the internal state $01$. If the actions $a[a]$
or $a[b]$ are chosen, they denote that the event type $a$ is currently
observed. The Markov decision process moves to the internal state $01$
and $11$ with probability $.375 = \frac{3}{8}$ and $.625 =
\frac{5}{8}$ respectively. Other transitions are defined similarly.

On the other hand, the initialization state $\underline{I}$
corresponds to the computation on neighbors. The Markov decision
process moves to the interal states $\underline{00}$,
$\underline{01}$, $\underline{10}$, $\underline{11}$ randomly with
the uniform distribution. Consider, for example, the internal state
$\underline{10}$. If the action $a[a]$ or $b[a]$ are chosen, they
denote that the current event type is replaced with the eveny type
$a$. The Markov decision process moves to the internal state
$\underline{00}$ and $\underline{10}$ with probability $.375 =
\frac{3}{8}$ and $.625 = \frac{5}{8}$ respectively. At an internal
state, the Markov decision process can move to the output states $0$,
$1$, or $2$ by the truncated $\frac{1}{2}$-geometric mechanism over
$\underline{2}$ (Figure~\ref{figure:randomized-frequency}, \`{a} la
Figure~\ref{figure:geometric-mechanism}). For any action sequence,
observe that paths from the state $I$ and those from $\underline{I}$
formalize the computation of Algorithm~\ref{algorithm:frequency-estimator} 
on an observation and its neighbor represented by the sequence. We
hence say the states $I$ and $\underline{I}$ are neighbors in the
Markov decision process (Figure~\ref{figure:frequency-estimator}).

@@

      (01) edge [loop below] node { $a[\uscore]$, .375 } (01)
      (01) edge [bend left=10] node { $a[\uscore]$, .625 } (11)
      (01) edge [loop left] node [above=1,rotate=90] { $b[\uscore]$, .625 } (01)
      (01) edge [bend left=10] node [left=-3,above=1,rotate=90] { $b[\uscore]$, .375 } (00)

@@

      (01') edge [loop below] node { $\uscore[a]$, .375 } (01')
      (01') edge [bend left=10] node { $\uscore[a]$, .625 } (11')
      (01') edge [loop left] node [above=1,rotate=90] { $\uscore[b]$, .625 } (01')
      (01') edge [bend left=10] node [left=-3,above=1,rotate=90] { $\uscore[b]$, .375 } (00')

@@

\noindent
\emph{Privacy Specification}.
For classical differential privacy, we are interested in the output
distribution. Algorithm~\ref{algorithm:frequency-estimator} has three
different outputs. By privacy analysis, the state $I$ satisfies the
following $\dpCTLstar$ formula:
\[
\dpriv{1}{0} \F 0 \wedge \dpriv{1}{0} \F 1 \wedge \dpriv{1}{0} \F 2
\]
For pan privacy, we moreover wish to measure privacy loss in internal
states. For observations of length $k$, the probabilities of reaching
each internal states need to be sufficiently close. The state $I$
therefore must satisfy
\[
\dpriv{1}{0} \F 0 \wedge \dpriv{1}{0} \F 1 \wedge \dpriv{1}{0} \F 2
\wedge
\bigwedge_{k \in \bbfN}
\dpriv{2}{0} \X^k 00 \wedge 
\dpriv{2}{0} \X^k 01 \wedge 
\dpriv{2}{0} \X^k 10 \wedge 
\dpriv{2}{0} \X^k 11.
\]
